Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-28 paper.tex
more edits to section 3, still in progress.  compactifying a bit with some clarification on how to derive.  will let you know when done with this round.

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, while we do not include viscosity in our
idealized disk models, in reality small-scale perturbations will be 
prone to viscous damping, and should drive weaker transport.

>>>>

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid analysis does not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

### 2015-04-28 setup.tex
more edits to section 3, still in progress.  compactifying a bit with some clarification on how to derive.  will let you know when done with this round.

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional 
disk orbiting a central star of mass $M_*$.  Cylindrical
co-ordinates $(r,\phi, z)$ are centered on the star. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\label{full_mass}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq.\ \ref{full_energy}) the sink term
$\Lambda$ includes non-adiabatic cooling and (negative) heating. 
The gas temperature $T$ obeys the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \OmK t_c
\end{align}
relative to the Keplerian frequency, $\OmK=\sqrt{GM_*/r^3}$.  We shall
use the terms `thermal relaxation' and `cooling' interchangeably
throughout this paper.

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, in \S\ref{application} 
we will consider a more realistic model for $\beta$ in
PPDs with a dust opacity (which gives Fig.  \ref{bcrit_mmsn2d}).

@@

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} 
These equations allows us to choose any midplane density 
distribution $\rho_0(r)\equiv \rho(r,0)$. In this work we assume a
power-law profile, 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index.

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives  a 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the temperature power-law index, and
define   
\begin{align}
   h(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\OmK$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $ h$
is also a characteristic disk aspect-ratio. We are 
interested in thin disks with $ h \ll 1$.%,%  as relevant for
% protoplanetary disks \citep{chiang97}.

%\subsection{Vertical structure}

Solving the equilibrium disk equations, we find the explicit
expression for the density field for 
$\Gamma\neq1$ is given  by
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
such that $\rho(r,H_s)=0$. 
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right) h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density and rotation profiles are
given, respectivley, by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
  \frac{\Omega^2(r,z)}{\OmK^2(r)}& =1+ (p+q) h^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for passively irradiated
PPDs \citep{chiang97}.

>>>>

\section{Governing equations and disk models}\label{setup}
An inviscid, non-self-gravitating disk orbiting a central star of mass $M_*$
obeys the three-dimensional fluid equations:
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\label{full_mass}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star with $G$ as the gravitational constant. The cylindrical
co-ordinates $(r,\phi, z)$ are centered on the star. 
In the energy equation (Eq.\ \ref{full_energy}) the sink term
$\Lambda$ includes non-adiabatic cooling and (negative) heating. 
The gas temperature $T$ obeys the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \OmK t_c
\end{align}
relative to the Keplerian frequency, $\OmK=\sqrt{GM_*/r^3}$.  We 
use the terms `thermal relaxation' and `cooling' interchangeably
throughout this paper.

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, in \S\ref{application} 
we will consider a more realistic model for $\beta$ in
PPDs with a dust opacity (as in Fig.\  \ref{bcrit_mmsn2d}).

@@

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{subequations}\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} \end{subequations}
A unique solution requires additional assumptions about disk structure.  We choose
\begin{subequations}\begin{align}
 \rho(r,0) &\equiv \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p, \\
 T(r,0) &\equiv T_0(r) = T_{00}\left(\frac{r}{r_0}\right)^q, \\
 P(r,z) & = K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align} \end{subequations}
where $r_0$ is a fiducial radius and $p$ and $q$ are the standard powerlaw indices
for midplane density and temperature, respectively.  The polytropic index $\Gamma$ 
parametrizes the vertical stratification with $\Gamma = 1$ ($\Gamma = \gamma$)
describing vertically isothermal (adiabatic) disks, respectively.  The ideal gas
law requires $K = \rho_0^{1-\Gamma}\mathcal{R} T_0/\mu$.

%These equations allows us to choose any midplane density 
%distribution $\rho_0(r)\equiv \rho(r,0)$. In this work we assume a
%power-law profile, 
%\begin{align}
%  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
%\end{align}
%where $\rho_{00}$ is the midplane density at the fiducial radius
%$r=r_0$ and $p$ is the power-law index.

%We consider disk equilibria in which the pressure $P$ and density
%$\rho$ are related by   
%\begin{align}
%  P(r,z) &= 
%  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
%  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
%\end{align}
%where the polytropic index $\Gamma$ is a  constant parameter  and the
%function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives  a 
%locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
%sound-speed. 
%It shall be convenient to define the function $c_s(r,z)$

We further define a modified sound speed
\begin{align}
	c_s \equiv \sqrt{\Gamma P /\rho } , 
% c_s^2\equiv % \frac{dP}{d\rho} =
%  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
%  \frac{\Gamma P}{\rho},  
\end{align}
which in general differs from the  isothermal, $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$, and adiabatic $c_\mathrm{ad} =
c_s\sqrt{\gamma/\Gamma}$, sound speeds.

We also introduce the characteristic scale-height, $H = c_s(r,0)/\OmK$
and disk aspect-ratio
%We take $c_{s0}(r)$ to be another power-law:  
%\begin{align}
%  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
%\end{align}
%where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the temperature power-law index, and
%define   
\begin{align}
   h(r) \equiv %\frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} \propto r^{(1+q)/2}
\end{align}
We are interested in thin disks with $ h \ll 1$.

\subsubsection{Density structure}
The equilibrium density field follows from vertical hydrostatic equilibrium (Eq.\ \ref{vert_equilibrium}).
The solution for $\Gamma\neq1$ is
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
where $\rho(r,H_s)=0$. 
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

The vertically isothermal case, $\Gamma = 1$, can be found either 
independently or as a limit of Eq. \ref{eqm_dens} to be 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \approx \exp {\left[ - {z^2 \over 2 H^2}\right]} \label{rhoisothin},
  \end{align}
  where Eq.\ \ref{rhoisothin} takes the $z \ll r$ thin disk limit.

 \subsubsection{Rotation and vertical shear profiles}
The equilbrium rotation field follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \left(\frac{p+q}{\Gamma}\right) h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\gamma$, where $s\equiv q+p(1-\Gamma)$.

Notice that for $\Gamma = 1$, $s/\Gamma = q$...

...
%\subsubsection{Vertically isothermal disks}
%For $\Gamma=1$ the equilibrium density and rotation profiles are
%given, respectivley, by 
%\begin{align}
%  \frac{\rho(r,z)}{\rho_0(r)} &=
%  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
%  \frac{\Omega^2(r,z)}{\OmK^2(r)}& =1+ (p+q) h^2(r) + q\left(1 -
%    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
%\end{align}
%which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
%taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
%vertically isothermal disks which is the relevant case for passively irradiated
%PPDs \citep{chiang97}.
